Efficient static analysis of Marlowe contracts

### 2020-04-30 papers/static_analysis2020/paper.tex
Add abstract and fix layout to keep it 16 pages

\maketitle

\section{Introduction\label{sec:Introduction}}

@@

State of the art libraries like SBV \cite{erkok2019sbv} allow straightforward
static analysis of arbitrary Haskell functions, with very few restrictions
on how those functions are implemented.

However, static analysis can be very computationally expensive, because
it essentially needs to check every possible execution of the contract.

@@

The case study presented is that of the optimisation of Marlowe\textquoteright s
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because it ran out of
memory after half an hour.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item Removing unnecessary parts from the analysis. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item Avoiding high level abstractions. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item Reducing search space by normalising parameters. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, then we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the fundamental structure of the implementation. We consider
two main ideas:
\begin{itemize}
\item Reducing search space by using normalised execution paths relevant
to the property. Instead of using search space to model inputs, we
use it to model possible executions, and we also ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item Minimizing the representation of inputs and outputs. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the most heavyweight optimisations described in the previous section.

@@

For example, in the case of Marlowe\textquoteright s static analysis,
Marlowe allows several inputs to be combined in one single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.

>>>>

\maketitle
\begin{abstract}
SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways that properties can be described. However, the way in which properties are described can affect the computational cost of verifying them considerably. In addition, it is quite hard to predict how different changes will affect the computational cost.

In this paper, we discuss the lessons we have learned while implementing and optimising the static analysis functionality for Marlowe, a domain specific language for describing self-enforcing financial smart-contracts that can be deployed on a blockchain.
\end{abstract}
\section{Introduction\label{sec:Introduction}}

@@

State of the art libraries like SBV \cite{erkok2019sbv} allow straightforward
static analysis of arbitrary Haskell functions, with very few restrictions
on how those functions are implemented. However, static analysis can be very
computationally expensive, because it essentially needs to check every
possible execution of the contract.

@@

The case study presented is that of the optimisation of Marlowe's
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because it ran out of
memory after half an hour.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item Removing unnecessary parts from the analysis. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item Avoiding high level abstractions. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item Reducing search space by normalising parameters. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the fundamental structure of the implementation. We consider
two main ideas:
\begin{itemize}
\item Reducing search space by using normalised execution paths relevant
to the property. Instead of using search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item Minimizing the representation of inputs and outputs. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the most heavyweight optimisations described in the previous section.

@@

For example, in the case of Marlowe's static analysis,
Marlowe allows several inputs to be combined in one single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.
